A Tidy Framework and Infrastructure to Systematically Assemble Spatio-temporal Indexes from Multivariate Data

### 2023-08-03 tidyindex.qmd
first two paragraphs in Sec 2

*Think about if there is another word for data pipeline*

**Why you should care about pipeline**

Data pipeline is not a new concept to computing. It refers to a set of data processing elements connected in series, where the output of one element is the input of the next one.
@wickham_plumbing_2009 argues that whether made explicit or not, the pipeline has to be presented in every graphics program.
The paper also argues that breaking down graphic rendering into steps is beneficial for understanding the implementation and comparing between different graphic systems. 
The discussion on pipeline construction is well documented in early interactive graphics software: @buja_elements_1988, @sutherland_orca_2000, and @xie_reactive_2014 and their pipeline steps include non-linear transformation, variable standardization, randomization and dimension reduction.

**What is pipeline, its underlying software design philosophy, and how these are reflected in R**

One of the most commonly known pipeline examples is perhaps the Unix pipeline where programs can be concatenated with `|` to flow the output from the last program into the next program, i.e.

````
command 1 | command 2 | command 3 | ...
````

To solve a complex problem, the Unix system builds simple programs that do one thing well and work well together.
This design is also reflected in the tidyverse ecosystem in R.
To solve a complicated data problem using tidyverse, analysts typically build the solution using a collection of tools from the tidyverse toolbox. 
The data object can flow smoothly from one command to the next, safeguarded by the tidy data format [@wickham_tidy_2014], which prescribes three rules on how to lay out tabular data.
The tidyverse tools also embrace a strong human-centered design where function names are intuitive and easy to reference through autocomplete.
With the tidyverse design principle in mind, the tidymodel suite enables analysts to build machine learning models through the data pipeline. 
It includes typical tasks required in machine learning like data resampling, feature engineering, model fitting, model tuning, and model evaluation.
An advantage of tidymodel pipeline over separate software for individual models is that analysts no longer need to write model-specific syntax to work with each model, but pipeline-specific syntax that is applicable to all the models implemented in tidymodel.
This allows users to easily experiment with a collection of machine learning models.

>>>>

<!-- **Why you should care about pipeline** -->

Data pipeline refers to data transforming through a series of functions, resembling the concept of function composition in mathematics. To accomplish specific tasks, analysts initially write lines of code and later generalise them into functions. This generalization transforms the code from a specific use-case to something reuseable and easy to control with arguments. A pipeline represents a further generalization of functions into a workflow. It chains individual functions together, rather than creating a single monolithic pieces, to solve a data analysis problem. The resulting pipeline can be reused for other similar problems, and ideally, allows for easy switching of functions within the pipeline, similar to controlling function parameters through arguments.

<!-- One of the most commonly known pipeline examples is perhaps the Unix pipeline where programs can be concatenated with `|` to flow the output from the last program into the next program, i.e. ````command 1 | command 2 | command 3 | ...````. To solve a complex problem, the Unix system builds simple programs that do one thing well and work well together. -->

<!-- **What is pipeline, its underlying software design philosophy, and how these are reflected in R** -->

Early discussion on data pipeline in R can be found in the interactive graphics literature [@buja_elements_1988; @sutherland_orca_2000; @xie_reactive_2014]. 
Specifically, @wickham_plumbing_2009 differentiates two different approaches used for structuring for (graphic) software: pipes and pipelines. The pipe approach offers functions, or components, and analysts concatenate functions together to form a pipeline to solve the specific problem at hands. This approach is used by the tidyverse suite, specifically in the dplyr package, where analysts apply functions in a sequence to address a wide array of data wrangling tasks. On the other hand, the pipeline approach addresses the data analysis tasks in a more structured manner, with a few centred steps used in almost all the workflow. For example, a supervised machine learning workflow typically involves steps like data splitting, model training and model evaluation, among others, regardless of the specific model used. A pipeline approach, like the tidymodel suite, modularizes these components and provides an unified syntax to specify different models.

### 2023-08-03 tidyindex.tex
first two paragraphs in Sec 2

\thanks{CONTACT: H. Sherry
Zhang. Email: \href{mailto:huize.zhang@monash.edu}{\nolinkurl{huize.zhang@monash.edu}}. }
\begin{document}
\captionsetup{labelsep=space}
\maketitle
\textsuperscript{1} Department of Econometrics and Business
Statistics, Monash University, Melbourne,
Victoria, Australia\\ \textsuperscript{2} Institute of
Statistics, University of Natural Resources and Life
Sciences, Vienna, Austria\\ \textsuperscript{3} Department of
Mathematical Sciences, BNU-HKBU United International College, Zhuhai,
Guangdong, China
\begin{abstract}
\begin{itemize}
\tightlist
\item
  indexes, useful, quantify severity, early monitoring,
\item
  A huge number of indexes have been proposed by domain experts,
  however, a large majority of them are not being adopted, reused, and
  compared in research or in practice.
\item
  One of the reasons for this is the plenty of indexes are quite complex
  and there is no obvious easy-to-use implementation to apply them to
  user's data.
\item
  The paper describes a general pipeline framework to construct indexes
  from spatio-temporal data,
\item
  This allows all the indexes to be constructed through a uniform data
  pipeline and different indexes to vary on the details of each step in
  the data pipeline and their orders.
\item
  The pipeline proposed aim to smooth the workflow of index construction
  through breaking down the complicated steps proposed by various
  indexes into small building blocks shared by most of the indexes.
\item
  The framework will be demonstrated with drought indexes as examples,
  but appliable in general to environmental indexes constructed from
  multivariate spatio-temporal data
\end{itemize}
\end{abstract}
\begin{keywords}
\def\sep{;\ }
indexes\sep data pipeline\sep 
software design
\end{keywords}
\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[boxrule=0pt, sharp corners, enhanced, breakable, borderline west={3pt}{0pt}{shadecolor}, frame hidden, interior hidden]}{\end{tcolorbox}}\fi

@@

\emph{Think about if there is another word for data pipeline}

\textbf{Why you should care about pipeline}

Data pipeline is not a new concept to computing. It refers to a set of
data processing elements connected in series, where the output of one
element is the input of the next one. Wickham et al. (2009) argues that
whether made explicit or not, the pipeline has to be presented in every
graphics program. The paper also argues that breaking down graphic
rendering into steps is beneficial for understanding the implementation
and comparing between different graphic systems. The discussion on
pipeline construction is well documented in early interactive graphics
software: Buja et al. (1988), Sutherland et al. (2000), and Xie,
Hofmann, and Cheng (2014) and their pipeline steps include non-linear
transformation, variable standardization, randomization and dimension
reduction.

\textbf{What is pipeline, its underlying software design philosophy, and
how these are reflected in R}

One of the most commonly known pipeline examples is perhaps the Unix
pipeline where programs can be concatenated with \texttt{\textbar{}} to
flow the output from the last program into the next program, i.e.~

\begin{verbatim}
command 1 | command 2 | command 3 | ...
\end{verbatim}

To solve a complex problem, the Unix system builds simple programs that
do one thing well and work well together. This design is also reflected
in the tidyverse ecosystem in R. To solve a complicated data problem
using tidyverse, analysts typically build the solution using a
collection of tools from the tidyverse toolbox. The data object can flow
smoothly from one command to the next, safeguarded by the tidy data
format (Wickham 2014), which prescribes three rules on how to lay out
tabular data. The tidyverse tools also embrace a strong human-centered
design where function names are intuitive and easy to reference through
autocomplete. With the tidyverse design principle in mind, the tidymodel
suite enables analysts to build machine learning models through the data
pipeline. It includes typical tasks required in machine learning like
data resampling, feature engineering, model fitting, model tuning, and
model evaluation. An advantage of tidymodel pipeline over separate
software for individual models is that analysts no longer need to write
model-specific syntax to work with each model, but pipeline-specific
syntax that is applicable to all the models implemented in tidymodel.
This allows users to easily experiment with a collection of machine
learning models.

@@

\leavevmode\vadjust pre{\hypertarget{ref-wickham_tidy_2014}{}}%
Wickham, Hadley. 2014. {``Tidy {Data}.''} \emph{Journal of Statistical
Software} 59 (September): 1--23.
\url{https://doi.org/10.18637/jss.v059.i10}.

>>>>

\thanks{CONTACT: H. Sherry
Zhang. Email: \href{mailto:huize.zhang@monash.edu}{\nolinkurl{huize.zhang@monash.edu}}. }
\begin{document}
\captionsetup{labelsep=space}
\maketitle
\textsuperscript{1} Department of Econometrics and Business
Statistics, Monash University, Melbourne,
Victoria, Australia\\ \textsuperscript{2} Institute of
Statistics, University of Natural Resources and Life
Sciences, Vienna, Austria\\ \textsuperscript{3} Department of
Mathematical Sciences, BNU-HKBU United International College, Zhuhai,
Guangdong, China
\begin{abstract}
\begin{itemize}
\tightlist
\item
  indexes, useful, quantify severity, early monitoring,
\item
  A huge number of indexes have been proposed by domain experts,
  however, a large majority of them are not being adopted, reused, and
  compared in research or in practice.
\item
  One of the reasons for this is the plenty of indexes are quite complex
  and there is no obvious easy-to-use implementation to apply them to
  user's data.
\item
  The paper describes a general pipeline framework to construct indexes
  from spatio-temporal data,
\item
  This allows all the indexes to be constructed through a uniform data
  pipeline and different indexes to vary on the details of each step in
  the data pipeline and their orders.
\item
  The pipeline proposed aim to smooth the workflow of index construction
  through breaking down the complicated steps proposed by various
  indexes into small building blocks shared by most of the indexes.
\item
  The framework will be demonstrated with drought indexes as examples,
  but appliable in general to environmental indexes constructed from
  multivariate spatio-temporal data
\end{itemize}
\end{abstract}
\begin{keywords}
\def\sep{;\ }
indexes\sep data pipeline\sep 
software design
\end{keywords}
\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[frame hidden, borderline west={3pt}{0pt}{shadecolor}, breakable, boxrule=0pt, enhanced, interior hidden, sharp corners]}{\end{tcolorbox}}\fi

@@

Data pipeline refers to data transforming through a series of functions,
resembling the concept of function composition in mathematics. To
accomplish specific tasks, analysts initially write lines of code and
later generalise them into functions. This generalization transforms the
code from a specific use-case to something reuseable and easy to control
with arguments. A pipeline represents a further generalization of
functions into a workflow. It chains individual functions together,
rather than creating a single monolithic pieces, to solve a data
analysis problem. The resulting pipeline can be reused for other similar
problems, and ideally, allows for easy switching of functions within the
pipeline, similar to controlling function parameters through arguments.

Early discussion on data pipeline in R can be found in the interactive
graphics literature (Buja et al. 1988; Sutherland et al. 2000; Xie,
Hofmann, and Cheng 2014). Specifically, Wickham et al. (2009)
differentiates two different approaches used for structuring for
(graphic) software: pipes and pipelines. The pipe approach offers
functions, or components, and analysts concatenate functions together to
form a pipeline to solve the specific problem at hands. This approach is
used by the tidyverse suite, specifically in the dplyr package, where
analysts apply functions in a sequence to address a wide array of data
wrangling tasks. On the other hand, the pipeline approach addresses the
data analysis tasks in a more structured manner, with a few centred
steps used in almost all the workflow. For example, a supervised machine
learning workflow typically involves steps like data splitting, model
training and model evaluation, among others, regardless of the specific
model used. A pipeline approach, like the tidymodel suite, modularizes
these components and provides an unified syntax to specify different
models.

@@
